A* with Perfect Potentials

### 2019-12-11 paper/ch_potentials.tex
capitalize title
prepare for arxiv
write stuff

\title{A* with perfect potentials} %TODO Please add

@@

\author{Ben Strasser}{Germany}{academia@ben-strasser.net}{}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hideLIPIcs

@@

\begin{abstract}
Quickly determining shortest paths in networks is an important ingredient for many routing applications.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical problems.
A* is an extension to Dijkstra's algorithm.
It uses a potential function to estimate the distance of each node to the target.
By adding these estimates to the queue keys, the search is directed towards the target.
The quality of this potential determines the performance of A*.
We introduce a novel way to efficiently calculate perfect potentials for extended problem settings where a lower bound graph is available.
For example, in the case of routing with live traffic, this could be the free flow graph.
\end{abstract}

\section{Introduction}
\label{sec:intro}

>>>>

\title{A* with Perfect Potentials} %TODO Please add

@@

\author{Ben Strasser}{Germany}{academia@ben-strasser.net}{TODO}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \EventEditors{John Q. Open and Joan R. Access}
% \EventNoEds{2}
% \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
% \EventShortTitle{CVIT 2016}
% \EventAcronym{CVIT}
% \EventYear{2016}
% \EventDate{December 24--27, 2016}
% \EventLocation{Little Whinging, United Kingdom}
% \EventLogo{}
% \SeriesVolume{42}
% \ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hideLIPIcs
\nolinenumbers

@@

\begin{abstract}
Efficiently computing flexible and optimal routes in continental-sized road networks is a thought challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical problems.
A* is the classical approach to accelerate Dijkstra's algorithm used in many fields such as Experimental Algorithms, Robotics, or Artificial Intelligence.
However, A*'s performance depends on the availability of a good potential, also called heuristic, function.
Computing a good potential is a challenge of its own.
In the past decade, many hierachical speedup techniques have therefore been developped.
They achieve speed and optimality but sacrifize flexibility.
In this paper, we use hierarchical techniques to compute potential functions.
With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
Efficiently computing flexible and optimal routes is a thought challenge.
Optimal means that the computed path always has minimum cost.
Being able to change the edge weights is flexibility.
Efficiently means that the path computation is fast.

Dijkstra's algorithm \cite{?} achieves flexiblity and optimality but unfortunately runs for several seconds on continental road networks, which is too slow for many applications.
A classic speedup approach is the well-known A* algorithm \cite{hnr-afbhd-68}.
While flexible, A*'s performance crucially depends on the quality of the potential or heuristic function.
Finding such a function is difficult.
A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{CH,CCH,CRP,HIGHWAYHIERACHIES,MLD} which do not require such a function.
A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{?} and its Live-Traffic capable Customizable CH (CCH) \cite{?} extension.
While undoubtedly fast, hierarchical approaches sacrifice flexibility to obtain speed.
In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.

Our algorithm is built upon ideas from Phast \cite{?}, a CH-extension.
The novel idea is to compute optimal A* potentials lazily.
The hierarchical approach is applied to the lower bound graph with respect to the edge costs.
As the hierarchical approaches are optimal, the resulting A* potential is the tightest possible.
With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.

TODO: Cite Sabine

Our technique can be applied to a CH and a CCH.
We describe our work mostly in CH terms.
Fortunately, all results are also directly applicable to CCH.
We describe the trade-off between the two variants in Section \ref{TODO}.

\section{Applications - or why do we care about Flexibility?}

\section{Outline}

\section{Flexible Two-Phase Problem Formulation}

Route planning algorithms usually operate in two or more phases.
The setup details vary among papers as different problem formulations are considered.
In this section, we describe the setup used in our work.

We consider two phases: a preprocessing and a query phase.
In a slow, offline \emph{preprocessing} phase, a graph $G$ with nodes $V$ and edges $E$ is given with \emph{lower bound edge weights} $w_\ell : E\rightarrow \mathbb{R}^+$.
In this phase, auxilary data is computed which is used in a second, fast query phase.

The input to the \emph{query phase} is a start or \emph{source node} $s$ and a destination or \emph{target node} $t$.
Further, the input to the preprocessing phase and the computed auxilary data is available.
Finally, a \emph{query weight function} $w:E\rightarrow \mathbb{R}^+$ is part of the query input.
It must hold that $w_\ell(e) \le w(e)$ for all edges $e$.
The output of the query phase is a path such that the sum over all $w$-weights is minimum.

Variations of this setup exist.
A very common assumption is that the query edge weights $w$ are known during the preprocessing phase.
This allows for very fast query running times, however, changing $w$ requires rerunning the slow preprocessing phase.
This setup is therefore very unflexible.
Many hierarchical speedup techniques such as CH use this setup.
Notable exceptions are \cite{MLD,CRP} and \cite{CCH} which use a third phase to change the query edge weights.

\section{A*}

A* star is a well-known extension to Dijkstra's algorithm.
It uses a potential $\pi:V\rightarrow \mathbb{R}^+$.
The potential is often called heuristic.
A* is optimal, if the potential $\pi(v)$ is consistent \cite{Pearl, Judea (1984). Heuristics: Intelligent Search Strategies for Computer Problem Solving. Addison-Wesley. ISBN 0-201-05594-5.}, which essentially mean that $\pi(v)$ is a lower bound of the distance from $v$ to $t$.

The best possible potential that can be derived from the auxilary data is the minimum path distance with respect to $w_\ell$.
Every better, optimal potential requires knowledge of the query weights $w$.
Our technique allows to compute these best possible potentials.
Within this regard, CH-potentials are optimal.

\section{CH}

\begin{algorithm2e}
\KwData{$B[x]$ tentative distance from $x$ to $t$, initially $+\infty$}
\KwData{Mininum priority queue $Q$, also called open list, initially empty}
Add $t$ to $Q$ with weight 0\;
$B[t] \leftarrow 0$\;
\While{not $Q$ empty}{
	$y\leftarrow$ pop minimum element from $Q$\;
	\For{$xy$ is down-edge in $G^+$}{
		\If{$B[x] > w_\ell(xy) + B[y]$}{
			$B[x]\leftarrow w_\ell(xy) + B[y]$\;
			\eIf{not $x$ in $Q$}{
				Add $x$ to $Q$ with weight $w_\ell(xy) + B[y]$\;
			}{
				Decrease weight of $x$ in $Q$ to $w_\ell(xy) + B[y]$\;
			}
		}
	}
}
\caption{CH backward search}
\label{algo:ch-backward}
\end{algorithm2e}

A CH is a two phase technique to efficiently compute optimal, shortest paths.
For a detailed exposition, we refer the interested reader to \cite{CCH, CH}.
In this section, we give a high-level overview of the aspects relvant to this paper.

A CH places nodes into levels.
No edge must connect two nodes within one level.
Levels are ordered by ``importance''.
An edge goes up when it goes from a node in a lower level to higher level.
Down edges are defined analogously.
An up-down path is a path where only one node $v$ is more important than its neighbors.
$v$ is called the \emph{mid} node.

In the preprocessing phase, a CH adds shortcut edges to the input graph $G$ to obtain $G^+$.
The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path.
We may therefore restrict our search to up-down paths.
This is done using a bidirectional search.
The forward search starts from $s$ and only follows up-edges.
Similarily, the backward search starts at $d$ and only follows down-edges in reversed direction.
The two searches meet in the mid node.
Pseudo-code for the backward search is presented in Listing~\ref{algo:ch-backward}.
The forward search works analogously.

A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case.

Using a CH, we directly obtain a basic CH-potential implementation.
We compute a CH with the lower bound edge weights $w_\ell$ in the preprocessing phase.
In the query phase, we run a classical $A^*$ search and execute in each potential function evaluation a CH query.

This works, however, while one CH query is fast, answering one for every node in the $A^*$ is slow.
Fortunatly, we can do better.
However, for this we need another component called Phast \cite{Phast}.

\section{Phast}

\begin{algorithm2e}
\KwData{$P[x]$ tentative distance from $x$ to $t$}
Execute Algorithm~\ref{algo:ch-backward}\;
\For{all CH levels $L$ from most to least important}{
	\For{all up edges $xy$ in $G^+$ with $x$ in $L$}{
		\If{$P[x] < P[y] + w_\ell(xy)$}{
			$P[x] \leftarrow P[y] + w_\ell(xy)$\;
		}
	}
}
\caption{Phast basic all-to-one search}
\label{algo:phast}
\end{algorithm2e}

Phast \cite{Phast} is a CH extension that computes distances from all nodes to one target node.
It works in two steps.
The first step is analogously to the CH query:
From the target node $t$, all reachable nodes via reversed down-edges are explored.
The second step iterates over all CH level from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all level are processed, the distances from all nodes towards $t$ is computed.
Phast is faster than Dijkstra's algorithm on road graphs because Phast is a better fit for modern processor architecutres.
We refer to \cite{Phast} for an in-depth experimental performance analysis.
Pseudo-code is provided in Listing~\ref{algo:phast}.

Using Phast, we can build another basic CH-potential version.
In the query phase, we first use Phast to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
In the next step, we run $A^*$ and implement the potential function as lookup in the array $H$.

This setup also works.
The potential evaluation and by extension the $A^*$ search is indeed fast.
However, Phast step before the search is comparatively expensive.
The reason for this is that the distance from all nodes towards $t$ are computed.
Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search towards $t$.

\section{CH-Potentials}

\begin{algorithm2e}
\KwData{$B[x]$ tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$P[x]$ memoized potential at $x$, $\bot$ if uncomputed}
\SetKwFunction{Pot}{Potential}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Pot{$x$}}{
	\If{$P[x] = \bot$}{
		$P[x]\leftarrow B[x]$\;
		\For{all up edges $xy$ in $G^+$}{
			\If{$P[x] > w_\ell(xy)+\Pot(y)$}{
				$P[x] \leftarrow w_\ell(xy)+\Pot(y)$\;
			}
		}
	}
	\Return{$P[x]$}\;
}
\caption{CH-Potentials Algorithm}
\label{algo:pot}
\end{algorithm2e}

Fortunately, the Phast computation can be done lazily using memoization as depicted in Listing~\ref{algo:pot}.
If a potential is unknown, it is computed by recursively computing the potentials of all higher nodes reachable via up-edges in $G^+$.
The computed potentials are then increased by the distance of the corresponding up-edge in $G^+$.
The potential of a node $x$ is the minimum over all these values and the distance found in the CH backward search.

\section{TopoCore}

TODO: Below is old
